---
title: "Introduction"
author: "Bryan Willauer"
date: "July 17, 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(farsfunctions)

```
## The Data

The data in this package comes from the National Highway Traffic Safety Administration (NHTSA) Fatality Analysis Reporting System (FARS) data.

```{rsample2013_data}
fars_2013_fn <- make_filename(2013)
fars_2013 <- fars_read(fars_2013_fn) 
dim(fars_2013)
fars_2013
```

For detailed information about the data, see the [NHTSA FARS Manuals & Documentation page](https://crashstats.nhtsa.dot.gov/#/DocumentTypeList/4). 

## Loading FARS Data

To load all of the data for a given year, use the `make_filename()` and `fars_read()` functions, as shown in the previous section.

### About the Filename

Use the `make_filename` command to create a filename to save/load new data.


```{r make_filename}
fars_2013_fn <- make_filename(2013)
fars_2013_fn
```

### Single Year

If you wish to just look at fatality data for a a single year, use the `fars_read_years()` function with a single year as input. The only data columns selected are `MONTH` and `year`.  This returns a list of length one, and the first element in the list is the `tbl_df` (the `tidyverse` data frame) listing the month and year for each fatal accident.

```{r single_year}
fars_2014 <- fars_read_years(years = 2014)
fars_2014[[1]]
```

### Multiple Years

If you wish to look at fatalities for multiple years, enter a vector of years as the argument for the `fars_read_years()` function (examples: `fars_read_years(years = c(2013, 2015))` or `fars_read_years(2013:2015)`.  Again, this returns a list of `tbl_df`s,  with each element of the list showing the month and year for each fatality. 

```{r multiple_years}
fars_3yrs <- fars_read_years(years = 2013:2015)
fars_3yrs
```

## Summarizing FARS Data

The `fars_summarize_years()` function take the same argument as the `fars_read_years()`, and produces a summary of the simple counts of fatalities by month and year: 

```{rsummarize_data}
fars_summary <- fars_summarize_years(2013)
fars_summary
```


## Mapping Fatal Crashes

Finally, the `fars_map_state` function takes a state ID number and a year, and maps that state's fatalities with a dot at the fatality location.    

For a list of the state ID numbers, see page 26 of the [FARS Analytical User's Guide (2015)](https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/812315#27). 

```{r mapping_crashes}

fars_map_state(10, 2014)
fars_map_state(36, 2014)
```
